Spatial and temporal variability in economic availability of residual biomass

### 2022-07-11 analysis/newms_elsevier.Rmd
first commit from squid

---
title: Spatial and temporal variability in economic availability of residual biomass
  from timber harvest in west-central Alberta, Canada (1990--2015)
author:
  - name: Glen W. Armstrong
    email: gwa@ualberta.ca
    affiliation: rr
    footnote: 1
  - name: Jay A. Anderson
    email: jaya@ualberta.ca
    affiliation: rees
  - name: Alex McPhee
    email: awmcphee@ualberta.ca
    affiliation: rees
address:
  - code: rr
    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
  - code: rees
    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  This is the abstract.

  It consists of two paragraphs.

journal: "Biomass & Bioenergy"
date: "`r Sys.Date()`"
bibliography: historicalResidueSupply.bib
linenumbers: true
numbersections: true
csl: elsevier-vancouver.csl

@@

```{r libraries, echo=FALSE, message=FALSE}
setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(scales)
library(gt)
library(quantmod)
library(bookdown)
```

@@

Humans have been using biofuels since we learned to control fire a
million or so years ago.  In developed economies, biofuels were
largely displaced by coal during the industrial revolution, and by
petroleum starting in the mid-nineteenth century.  In recent times,
biofuels are an increasingly important source of energy because of
their potential to reduce dependency on fossil fuels and reduce
greenhouse gas emissions.  Direct combustion of biomass is still an
important source of energy.  In Europe, in particular, many
electricity plants have been converted to use wood pellets instead of
coal as the primary energy source.  Liquid biofuels (*e.g.* ethanol
and biodiesel) have been produced in commercial quantities from food
crops starting in the 1980s.  These are called first generation
biofuels.  A major difficulty with these first generation biofuels is
that they compete with food and that their use  may lead to land use change.

Second generation biofuels rely on processes that convert
lignocellulosic materials (such a corn stover and wood) to liquid
biofuels.  An advantage of this process is that the feedstock can be
residue from food or forest products production.  Second generation
biofuel production is still largely in a pilot or demonstration phase.  An
interesting example using forest residues is the Silva Green Fuel
renewable crude oil plant in Tofte, Norway which is expected to begin
operations in 2021 @Lane2018.

Wood pellets are a well-established biofuel.  It is projected that
Canada will produce 3.8 million Mg of wood pellets in 2021, of which
87% will be exported, largely to the United Kingdom, Japan, and Europe
@Watters2021.  Wood pellets are replacing coal as the primary energy
source in electricity generating plants, particularly in Europe
@Sterman2018.  Most of the wood pellet production in Canada comes from
mill residues.  There is interest in expanding the source to the forest
by collecting the residue left on cutblocks after harvest: this
practice is well-established in Europe.  Forest
residues are generated as by-products of conventional forest
harvesting operations, and include the tree tops, branches and
non-merchantable logs. These residues are typically left behind in
slash piles, which are burned during winter to mitigate against the
risk of wildfires. Canada has a forestry sector that produces
approximately 20 million oven-dry metric tons of forest residues
annually @Dymond2010.  For the remainder of this paper we will express
biomass using SI units, and all quantities of biomass will be
expressed on an oven-dry basis: 1 oven-dry metric ton is equivalent to
1 Mg.

~~Clean fuel standard
https://gazette.gc.ca/rp-pr/p1/2020/2020-12-19/html/reg2-eng.html
Despite the potential to produce biofuels from forest residues, such
use is limited to a few highly subsidized pilot projects. Several
operational and economic challenges hinder the utilization of forest
residues. These challenges are related to: (1) capital investments in
biorefineries, (2) geographical distance between feedstock and
markets, and (3) concerns surrounding feedstock availability
@Shabani2013. The purpose of our study is to explore concerns
surrounding this third challenge -- feedstock availability -- by
evaluating annual variability in commercial stemwood harvest in the
province of Alberta, Canada.~~

A number of studies have shown that collection and transportation
costs may be the biggest impediment to widespread use of forest
residues in bioenergy systemsX
[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
for a bioenergy plant largely depends on commercial harvesting operations
and the production of traditional forest products. Fluctuations in the
volume of stemwood harvested for forest products has a direct
impact on residues available for biofuels. Major factors, such as the
2008 housing crisis and the ongoing softwood lumber dispute with the
United States have resulted in sawmill closures and reductions in
forest harvesting activities
@deAvillez2014 @Spelter2009. This relationship between volatile
harvest volumes and subsequent residue variability is explored in detail by
Niquidet and Friesen @Niquidet2014, who model forest residue
supply in Alberta as a  function of lumber prices. Given that
biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
costs @Stephen2010, reliable feedstock supplies
and costs over the life of the bioenergy plant are important factors to
consider when deciding where to locate the bioenergy plant.

Past studies regarding bioenergy plant locations have
improved our understanding of feedstock availability and constraint.
One key area that has largely been omitted is considerations of
feedstock supply variability over time. Commercial stemwood harvests
vary year-to-year, largely based on market conditions for forest
products. Since a forest-residue based bioenergy plant is dependent on
stemwood harvesting for feedstock, any variability in harvesting
directly influences the volume of residues that would be available.
Therefore, considering the variability associated with feedstocks in
different locations could be important in choosing the optimal site for
locating a bioenergy plant

**We say above that consideration of variable feedstock supplies has
been omiXtted, and in the paragraph below that a bunch of people have
consider it.  Do those studies actually consider variability?**

~~Some studies
have considered variable feedstock supplies when investigating whether
and when to expand the production of bioenergy from burning residues
 @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other
studies have explored the context of variable supplies when making
decisions about capital investments in forest residue-based
biorefineries  @Cambero2015, @Chen2012 @Papapostolou2011. Variable
feedstock supplies are also shown to affect capital investment
decisions related to: investing in a residue preprocessing facility
(Chen and Fay 2011) @Chen2012, types of storage infrastructure for
bioenergy plant feedstocks 
@Papapostolou2011, as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities  @Cambero2015.~~

~~There has been a number of studies conducted regarding siting decisions
for forest residue-based biorefineries -- see Johnson et al (2012) @Johnson2012  for
a review of these studies -- however,~~

We are aware of no study that
considers historical variability of stemwood harvest and its subsequent effect on
forest residue availability. Our study contributes to the
literature by considering variability in forest residue feedstock
supplies.

More specifically, we analyze how this variability affects the
cost of delivering residue, and then use these cost estimates to inform
the decision on where to locate a bioenergy plant.

We consider co-locating the bioenergy adjacent to existing forest
products manufacturing facilties. Co-locating with existing facilities
and making use of established infrastructure and transportation
systems can reduce many of the costs associated with establishing a
new bioenergy plant @Browne2012 @Saddler2012 @Benjamin2009 @Towers2007
(**I don't have the references for Browne2012, Saddler2012,
Benjamin2009, or Towers2007**).  Bioenergy plant location should be
informed by how stable the respective residue supplies are. More
stable stemwood harvest would generate a more stable supply of forest
residues that could be collected from the nearby harvested areas.

We use information on the location and year of timber harvest in the
province of Alberta, Canada to quantify the spatial and temporal
variability of timber harvest, and relate that information to spatial
and temportal variability of production of timber harvest residues.
We develop transportation cost estimates to the timber harvest areas
based on cycle time estimates developed from the road network in the
province. From this information we develop historical techno-economic
supply curves [@Galik2009,@Yemshanov2014,@Blackburn2020, @Swinton2021]
for forest residue biomass for each of 26 years in our study period
(1990--2015) for 3 potential biofuel plant locations.

~~We use information The objectives of our study are ...... Our approach
incorporates variable feedstock availability into decision making, and
our results provide important information for assessing the potential
for a forest residue-based biofuel industry in Alberta.~~

The methods used for this study are similar to that used by
FPInnovations for their biomass availability estimations for timber
supply areas in British Columbia (see Ref. @Blackburn2020 for an
example).  Like us, FPInnovations develops supply curves
(cost-availability curves) based on a road network and historical
harvests.  However, their take on temporal variability is limited to
examining 2 5-year periods. As well, they limit their analyses to the
boundaries of timber supply areas.  We do not limit ourselves with
arbitrary boundaries.

Our analysis also shares similarities with work published by
Yemshanov *et al.* @Yemshanov2014. They devolop supply curves at
national and regional levels based on a national forest inventory and
the location of cogeneration facilities associated with forest
products manufacturing facilities.

The main contribution of this paper is the consideration of multiple
years to develop estimates of the temporal variability in forest
residue supply.

~~OTHER POTENTIALLY USEFUL TIDBITS.. I left these here as a reminder.~~

~~The need to go outside its normal delivery zone in bad years is similar
to a scenario in the agriculture residue feedstock supply literature,
called the the "derisked" scenario, whereby a buffer area outside the
normal supply zone is contracted to provide residues during years when
residue supply is low @Golecha2016.~~

~~Modern biorefineries can expect to breakeven financially with
delivered biomass costs of approximately \$80 (US\$60) Mg (Steeper
2021) @Steeper2021 . Studies have shown that forest policy reforms
could reduce delivered forest residue costs by incenting plantations
of fast-growing tree species on lands close to pulpmill locations
(Shooshtarian et al. 2021 @Shooshtarian2021, Anderson et al. 2012
@Anderson2012). Other important factors will be advances in
biorefinery technology and increases in biofuel prices. In Canada,
governments impact biofuel prices by setting renewable fuel mandates
and paying subsidies for biofuel production (Campbell et al.,
2016). The continuation of these mandates and subsidies will likely be
an important factor for the emergence of a second-generation
("advanced") biofuel sector. Also, the Government of Canada\'s Clean
Fuel Standard, which is proposed to come into force in 2022
(Government of Canada, 2020), is expected to improve the economics of
advanced biofuel production.~~

# Materials and methods

We describe here our study area, data sources, and methods.

@@

The southeastern part of the province is largely privately owned
agricultural land, and the northern and western parts are largely
publicly owned forested land.  There is, however, a substantial area of
privately owned agricultural land in the Peace River Country of
northwestern Alberta.  There are also large areas in National
Parks along the Rocky Mountains and in the far north of the province.
The province calls the area of mostly publicly owned forested land
(excluding those National Parks) the Green Area, and the area of
mostly privately owned agricultural land the White Area.  About 90%
of the harvested timber volume in Alberta comes from the Green Area.

For administrative purposes, the Green Area is divided into Forest
Management Units (FMUs).  The province enters into long term
agreements with forest products companies giving the companies the
rights to timber harvest and the responsibility to manage the forest
sustainably.  These agreements are called Forest Management
Agreements, and the areas under agreement are called Forest Management
Agreement Areas (FMAAs).  For the most part, the boundaries of the
FMAAs are coincident with FMU boundaries, with the exception of
the Alberta-Pacific FMAA in eastern Alberta which comprises 12 FMUs.  Maps
of the FMUs @AlbertaFMU and the FMAAs @AlbertaFMA can be found on the
Government of Alberta website.  The FMU and FMAA boundaries are
important to this study as they are used to assign average piece size
and utilization standards.  We discuss piece sizes and utilization
standards later.

Our focus for this study is the forest near Hinton, Edson, and
Whitecourt in west-central Alberta.  These locations were chosen as
they host several forest products companies and are located near
forest areas with substantial logging activity.  We assume that a
residual forest biomass processing plant would be located near an
existing forest products mill as the forest road network has been
developed to supply those mills. Locating a forest residue processing
plant near a forest products mill would also allow for convenient
access to mill residues.

The FMAAs which supply most of the timber to mills  near our
three locations are West Fraser Mills Ltd. (Hinton), West Fraser Mills
Ltd. (Edson), Weyerhaeuser Company Limited (Pembina Timberlands),
Millar Western, Blue Ridge Lumber Inc., and ANC Timber Ltd.

We used the Weyerhaeuser oriented strand board plant in Edson, the
Millar Western Forest Products Ltd. pulp mill in Whitecourt, and the
West Fraser pulp mill in Hinton as the locations for the hypothetical
forest residue processing facilities.

As well as considering locations for the hypothetical plants, we
considered 3 different plant capacities defined by annual biomass
requirements: 200 Gg/a corresponding to a small wood pellet mill, 400
Gg/a corresponding to a large pellet mill or a small ethanol plant,
and 800 Gg/a corresponding to a large ethanol plant.

## Data sources

We used a number of publicly available datasets to create the
township-level summary data used in our analysis.  Most of the data
are freely available on the World Wide Web (WWW). Much of the data
processing required for this study was manipulation of spatial data.
This was done using the QGIS geographic information system software
@QGISorg2021 and add-ons.  All national level data were clipped to the
Alberta provincial boundary for further analysis.

@@

Altalis manages much of the spatial data created by or for the
Government of Alberta.  We use several coverages from their base
features database product @Altalis2021: boundaries for the province,
the Green and White Areas, forest management units (FMUs), forest
management agreement areas (FMAAs), Alberta Township Survey System
(ATS_ townships, and towns.

The ATS boundaries are particularly important to this study as all the
data used for this study are aggregated or disaggregated to the
township level.  ATS provides the basis for legal land descriptions in
Alberta.  The typical township is roughly square (9.778 km by 9.716
km, 9 500 ha) and is indexed by township (1--126), range (1--30), and
meridian (4--6).  The townships immediately east of Alberta's western
boundary, and immediately east of the 5^th^ and 6^th^ meridians
(corresponding to 114°W and 118°W longitude, respectively) are partial
townships and therefore deviate from the roughly square shape.  ATS is
described in detail by the Alberta Land Surveyor's Association
@ALSA2021.  There are 7 237 townships in Alberta, allowing for spatial
resolution appropriate for regional or provincial level analyses.

@@

Alberta Geospatial Services is operated by the Alberta Ministry of
Environment and Parks @Alberta2021a.  We obtained road access
coverages from there.  A road access coverage is also available from
Altalis but it does not have as detailed information on road class.

### National Forest Information System

The National Forest Information System (NFIS) provides a set
of spatial data layers representing forest attributes at a 250 m
resolution for all of Canada @Beaudoin2017.  We used the merchantable
volume, stand age, and species composition attributes from these
data layers.

Another product of NFIS is a set of spatial datasets derived from
satellite data @NFIS2021a.  The dataset used for this study shows the
location of timber harvest for each of the years 1985-2015.  The
dataset is described in detail in Ref.  @Hermosilla2016.  It consists
of a raster for Canada at 30 m resolution indicating the year of
harvest for each cell.

The NFIS also provides a web-based individual tree biomass calculator
@NFIS2021b.  It is used to calculate stem wood, stem bark, branch, and
foliage biomass for a tree based on province, ecozone, species,
diameter at breast height (DBH) and height.  It is possible to upload
a file representing many trees with different combinations of these
input parameters and download a file containing the calculated results
for all the trees.

@@

Ref.  @Ung2013 describes the tree taper models developed for Canada.
~~A valuable contribution related to this is the WWW-based wood volume
calculator that uses those models @NRCan2015.~~  We use these taper
models to determine merchantable volumes and the amount of residual
biomass available from cutblocks in each of the townships in each
year.

### National Forestry Database

The National Forestry Database @CCFM2021 was used to summarize the
timber harvest volume and harvest area from Alberta for the years
1990-2015. This information is used to adjust the volumes obtained
from the 250 m resolution inventory to better reflect average harvest
volumes observed in Alberta.

@@

The boundaries of Canada's terrestrial ecozones can be found on
Agriculture and Agri-Food Canada's website @Canada2017.  Ecozone is an
input to the individual tree biomass calculator @NFIS2021b.

@@

The GIS analysis for this paper was done using QGIS @QGISorg2021.
Data summaries and statistical analysis were done using R @R2021 with
RStudio @RStudioTeam2021, and tidyverse @Wickham2019.

The primary objective of this paper was to quantify the spatial and
temporal variability of the costs associated with recovering the
residual biomass left on forest cutblocks following timber harvest.  A
driving assumption for this model was that residue recovery is
opportunistic.  Loggers make the decision about what and where to
harvest, and the biomass recovery occurs at some time following timber
harvest.  The decision to harvest is made independently of any
knowledge of the value of residue as biomass.  The biomass
~~gleaners~~ harvesters
provide a service to the loggers as the residue would be piled and 
burned if it was not recovered as biomass.

Every township is  associated with a cycle time (and cost)
required to transport harvest residue from the township to each of our
3 plant locations, a measure of residual biomass for each of the 26
years in the study period. For each plant location and year, the model
examines all the townships in the province that have a positive
residual volume, and begins the creation of cumulative residual
biomass curves starting with the lowest cycle time first.  These lists
of cycle time and cumulative biomass provide the data for the marginal
cycle time and cost curves.

This is a greedy algorithm.  It ignores the cost of moving equipment
from one block to another, but we believe it still provides useful
information.

## Cycle time estimates

@@

$$
	\bar{s_h}=\left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}
$$

where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.  Use
of harmonic mean allows us to calculate the correct cycle time: the
time required to travel from the mill, to the collection site at the
cutblock, and back to the mill.  This provides the basis for our
estimates of haul cost.

Figure 2 illustrates the road network in Alberta and estimates of
cycle time based on the average of loaded and unloaded speeds
from Table 1. The road network in the White Area of the province is
very dense and largely arranged as a grid related to the Alberta
Township Survey System.  Most of the roads in the White Area are
public.  The road network in the Green Area is less dense and
many of the roads are built and maintained by resource industries
(*e.g.* forestry and petroleum & natural gas).  The pattern of the
road network is not as regular as in the White Area.

[ Figure 2 about here]

One-way haul times based on the harmonic mean travel speed were
generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
~~created by Raffler~~ @Raffler2018, specifically its iso-area as
interpolation (from point) algorithm.  In our case, we used the
algorithm to determine the minimum one-way travel time (using the
harmonic mean of loaded and unloaded speeds) along the road network
from each of our centres to each cell of a provincial level raster at
a 200 m resolution.  Off-road travel was assumed to be at 5 km/h.  The
Dijkstra algorithm @Dijkstra1959 is the basis for travel time
minimization.  Cycle times were calculated by multiplying the one-way
haul time by 2, and adding 1 hour for loading and unloading.

QGIS zonal statistics were used to calculate the median cycle
time to each township, from each of our three centres.  The 
cycle  time in Figure 2 was limited to 14 hours because, in Alberta,
drivers of commercial vehicles are not permitted to drive
more than 13 hours in a 24-hour period, and spend no more that 14
hours on-duty, including loading and unloading times.

```{r constants, echo=FALSE, include=FALSE}
#hourly_truck_rate <- 175 + 225 + 40 #place holder.
# info fromhttps://www.rogerstrucking.ca/services/rates-and-equipment/62-tractor-truck-hourly-rates
# 175 $/hr for tractor, 225 for walking floor trailer, 40 for operator

>>>>

---
title: Spatial and temporal variability in economic availability of residual biomass
  from timber harvest in west-central Alberta, Canada (1990--2015)
author:
  - name: Glen W. Armstrong
    email: gwa@ualberta.ca
#    affiliation: rr
    footnote: 1
  - name: Jay A. Anderson
    email: jaya@ualberta.ca
#    affiliation: rees
  - name: Alex McPhee
    email: awmcphee@ualberta.ca
#    affiliation: rees
#address:
#  - code: rr
#    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
#  - code: rees
#    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, #Alberta, Canada, T6G 2H1
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  We use historical forest harvest data from Alberta, Canada to develop
  a description of the spatial and temporal distributions of the
  availability of forest harvest residues for the period 1990---2015.
  Using a GIS coverage of the road network in Alberta, we developed
  estimates of round-trip haul costs from each harvest location to each of
  three potential bioenergy plant locations. This information was used to
  create supply curves (or cost-availability curves) for each of the three
  potential locations and each of the 26 years. This information is used
  to develop descriptions of the variability in the delivered costs of
  forest residue for bioenergy use. These descriptions of variability are
  historical, but could be useful for potential investors in the bioenergy   sector.

journal: "Biomass & Bioenergy"
date: "`r Sys.Date()`"
bibliography: [historicalResidueSupply.bib]
linenumbers: true
numbersections: true
csl: elsevier-without-titles.csl

@@

```{r libraries, echo=FALSE, message=FALSE}
setwd("H:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(scales)
library(gt)
library(quantmod)
library(bookdown)
```

@@

Humans have used biofuels since learning to control fire a million or so
years ago. In developed economies, early biofuels (*e.g*. wood,
charcoal, dung, peat) were largely displaced by coal during the
industrial revolution, and by petroleum starting in the mid-nineteenth
century. In recent times, biofuels are once again becoming an important
source of energy because of their potential to reduce greenhouse gas
emissions. In the European Union, the United Kingdom, and Japan, many
heat and power plants have been converted to use wood pellets instead of
coal or natural gas.

Liquid biofuels (*e.g.,* ethanol and biodiesel) have been produced in
commercial quantities from food crops starting in the 1980s. In some
jurisdictions policies to promote the use of these first-generation
biofuels are being phased out because of concerns that they compete with
food, and that their production may lead to land use change @Campbell2016.

Second generation biofuels rely on processes that convert
lignocellulosic materials (such a corn stover and wood) to liquid
biofuels. An advantage of this process is that the feedstock can be
waste residue left over from food or forest products production. Second
generation biofuel production is still largely in a pilot or
demonstration phase. An interesting example that will use forest
residues is the Silva Green Fuel renewable crude oil plant in Tofte,
Norway which was expected to begin operations in 2021 @Lane2018. As of the
date of writing (2022-07-08), the plant is not yet operational.

Technologies to pelletize wood are more established than technologies to
produce liquid fuels from wood. It was projected that Canada would
produce 3.8 million Mg of wood pellets in 2021, of which 87% will be
exported, largely to the United Kingdom, Japan, and Europe
@Watters2021.  Wood pellets are replacing coal as the primary energy
source in electricity generating plants, particularly in Europe
@Sterman2018.  Most of the wood pellet production in Canada comes from
mill residues.  There is interest in expanding the source to the forest
by collecting the residue left on cutblocks after harvest: this
practice is well-established in Europe.  Forest
residues are generated as by-products of conventional forest
harvesting operations, and include the tree tops, branches and
non-merchantable logs. These residues are typically left behind in
slash piles, which are burned during winter to mitigate against the
risk of wildfires. Canada has a forestry sector that produces
approximately 20 million oven-dry metric tons of forest residues
annually @Dymond2010.  For the remainder of this paper we will express
biomass using SI units, and all quantities of biomass will be
expressed on an oven-dry basis: 1 oven-dry metric ton is equivalent to
1 Mg.

A number of studies have shown that collection and transportation
costs may be the biggest impediment to widespread use of forest
residues in bioenergy systemsX
[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
for a bioenergy plant largely depends on commercial harvesting operations
and the production of traditional forest products. Fluctuations in the
volume of stemwood harvested for forest products has a direct
impact on residues available for biofuels. Major factors, such as the
2008 housing crisis and the ongoing softwood lumber dispute with the
United States have resulted in sawmill closures and reductions in
forest harvesting activities
@deAvillez2014 @Spelter2009. This relationship between volatile
harvest volumes and subsequent residue variability is explored in detail by
Niquidet and Friesen @Niquidet2014, who model forest residue
supply in Alberta as a  function of lumber prices. Given that
biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
costs (@Caputo2005 @Leistritz2007 cited in @Stephen2010),a number of studies have shown that
collection and transportation costs may be the biggest impediment to
widespread use of forest residues in bioenergy systems. reliable feedstock supplies.
Since a
forest-residue based bioenergy plant is dependent on stemwood harvesting
for feedstock, any variability in harvesting directly influences the
volume of residues that would be available. Therefore, considering the
long-term variability associated with feedstocks in different potential
locations could be important in choosing the optimal site for locating a
bioenergy plant.

and costs over the life of the bioenergy plant are important factors to
consider when deciding where to locate the bioenergy plant.

Past studies regarding bioenergy plant locations have
improved our understanding of feedstock availability and constraint.
One key area that has largely been omitted is considerations of
feedstock supply variability over time. Commercial stemwood harvests
vary year-to-year, largely based on market conditions for forest
products. Since a forest-residue based bioenergy plant is dependent on
stemwood harvesting for feedstock, any variability in harvesting
directly influences the volume of residues that would be available.
Therefore, considering the variability associated with feedstocks in
different locations could be important in choosing the optimal site for
locating a bioenergy plant.

Although few studies consider long-term forest residue variability for
the purposes of optimizing the location of a potential bioenergy plant,
there are studies that consider variability for other purposes. Some
studies have considered variable feedstock supplies when investigating
whether and when to expand the production of bioenergy from burning
residues @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other studies have explored the context
of variable supplies when making decisions about capital investments in
forest residue-based biorefineries @Cambero2015 @Chen2012 @Papapostolou2011. Variable
feedstock supplies are also shown to affect capital investment decisions
related to investing in a residue preprocessing facility (Chen and Fay
2011) Chen2012, types of storage infrastructure for bioenergy plant
feedstocks Papapostolou2011, as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities @Cambero2015.

Our study contributes to the literature by considering variability in
forest residue feedstock supplies for the purpose of selecting the best
location for a bioenergy plant from three potential locations. More
specifically, we use precise information on the location and year of
historical timber harvest in the western Canadian province of Alberta to
quantify the spatial and temporal variability of timber harvest, and
thereby relate the harvest information to spatial and temporal
variability of forest residue availability. We develop transportation
cost estimates to all of the historical timber harvest areas based on
cycle time estimates developed from the road network in the province.
The spatial and temporal variability of forest residues results in
variability in the transport cycle time from the residue collection
areas to the bioenergy plant, which in turn leads to variability in
delivered residue cost. From this information we develop historical cost
curves for collecting and delivering forest residues to each of
the three potential bioenergy plant locations for each of 26 years in
our study period (1990--2015). When considering locations for the
potential bioenergy plants, we conduct sensitivity analysis on 3
different bioenergy plant capacities, as defined by annual forest
residue feedstock requirements: 200 Gg/a corresponding to a small wood
pellet mill, 400 Gg/a corresponding to a large pellet mill or a small
biorefinery, and 800 Gg/a corresponding to a large biorefinery.

The methods used for this study are similar to that used by
FPInnovations for their biomass availability estimates for timber supply
areas in British Columbia (see Ref. @Blackburn2020 for an example). Like us,
FPInnovations develops supply curves (cost-availability curves) based on
a road network and historical harvests. However, their take on temporal
variability is limited to comparing two 5-year periods. As well, they
limit their analyses to the boundaries of timber supply areas. The only
boundary that limits our analysis is the provincial border between
Alberta and British Columbia.

Our analysis also shares similarities with work published by Yemshanov
*et al.* @Yemshanov2014. They develop supply curves at national and regional
levels based on a national forest inventory and the location of
cogeneration facilities associated with forest products manufacturing
facilities. However, their analysis is based the reported annual volumes
of processed wood at each manufacturing plant in a single year (2010),
and the corresponding residue delivery cost. Instead, we consider the
spatial location of all the harvest sites over a 26-year period
(1990-2015) and calculate the delivery cost for each harvest site in the
year it was harvested.

In the next section we describe our study area and data sources. We then
describe how we aggregate and analyze the data. We present our
distributional results in the form of marginal delivered biomass cost
for each of the 26 years in the study period, and for each of the three
alternative plant locations. We conclude with a discussion of financial
considerations regarding forest residues as a biofuel feedstock, with
relevance to forest companies, policy makers, and bioenergy investors.

# Study area and data

@@

The southeastern part of the province is largely privately owned
agricultural land, and the northern and western parts are largely
publicly owned forested land. There is, however, a substantial area of
privately owned agricultural land in the Peace River Country of
northwestern Alberta. There are also large areas in National Parks along
the Rocky Mountains and in the far north of the province. The provincial
government identifies the area of mostly publicly owned forested land
(excluding the National Parks) as the Green Area, and the area of mostly
privately owned agricultural land as the White Area. About 90% of the
harvested timber volume in Alberta comes from the Green Area.

Our focus for this study is area surrounding the towns of Hinton, Edson,
and Whitecourt in west-central Alberta. These 3 potential bioenergy
plant locations were chosen as they host forest products companies and
are located near forest areas with substantial harvesting activity. We
assume that a potential forest residue-based bioenergy plant would be
located near an existing forest products mill, as timber harvesting is
occurring to supply those mills, and the requisite forest road network
would already be developed. Locating a forest residue processing plant
near a forest products mill would also allow for convenient access to
mill residues.

We used the Weyerhaeuser Company Limited oriented strand board plant in
Edson, the Millar Western Forest Products Ltd. Pulp mill in Whitecourt,
and the West Fraser pulp mill in Hinton as the three locations that we
consider for a potential forest residue-based bioenergy plant.

## Data

We used several publicly available datasets to create the township-level
(typically 9.778 km by 9.716 km, or 9 500 ha) summary data used in our
analysis. Most of the data are freely available on the World Wide Web.
All national level data were clipped to the Alberta provincial boundary
for further analysis.

@@

Altalis manages much of the spatial data created by or for the
Government of Alberta. We use several coverages from their base features
database product @Altalis2021: boundaries for the province, the Green and
White Areas, forest management units (FMUs), forest management agreement
areas (FMAAs), and Alberta Township Survey System (townships provide the
basis for legal land descriptions in Alberta).

@@

Alberta Geospatial Services is operated by the Alberta Ministry of
Environment and Parks @Alberta2021. We obtained road access coverages from
this database.

### National Forest Information System (NFIS)

The NFIS provides a set of spatial data layers representing forest
properties at a 250 m (6.25 ha) resolution for all of Canada @Beaudoin2017. We
used the merchantable volume, stand age, and species composition
attributes from these data layers.

Another product of NFIS is a set of spatial datasets derived from
satellite data @NFIS2021a. The dataset used for this study shows the
location of timber harvest for each of the years 1985-2015. The dataset
is described in detail in Ref. @Hermosilla2016. It consists of a raster for
Canada at 30 m (0.09 ha) resolution indicating the year of harvest for
each cell.

The NFIS also provides a web-based individual tree biomass calculator
@NFIS2021b, which we used to calculate stem wood, stem bark, branch, and
foliage biomass for a tree based on province, ecozone, species, diameter
at breast height (DBH) and height.

@@

Ref. @Ung2013 describes the tree taper models developed for Canada. We use
these taper models to determine merchantable stemwood volumes and the
amount of forest residues (branches and tops) available from harvested
areas in each of the townships in each year.

### National Forestry Database (NFD)

The NFD @CCFM2021 provided the timber harvest volumes and total harvest
areas from Alberta for the years 1990-2015. We used this information to
adjust the spatial NFIS volumes mentioned above to better reflect the
actual aggregated harvest volumes observed in Alberta.

@@

The boundaries of Canada's terrestrial ecozones can be found on
Agriculture and Agri-Food Canada's website @Canada2017. Ecozone is an input
into the NFIS individual tree biomass calculator @NFIS2021b, described in
section 2.2.3.

@@

We quantify the spatial and temporal variability of the costs associated
with collecting and transporting the forest residues left following
stemwood harvest. The variability of costs is largely related to changes
in the transport distance from the residue collection areas to the
bioenergy plant.

Most of the effort expended on this study was spent in aggregating
township-level summaries of data for use by the model. Much of this data
processing was the manipulation of spatial data, which was done using
the QGIS geographic information system (GIS) software @QGISorg2021 and
add-ons. Data summaries and statistical analysis were done using R
@R2021 with RStudio @RStudioTeam2021, and tidyverse @RStudioTeam2021.

We develop a model based on the assumption that forest residue
extractors follow a greedy optimization procedure: they take the
cheapest (closest) residue first, and stop when they have enough to
satisfy the annual biomass requirements for the three scales of
bioenergy plants: 200 Gg/a, 400 Gg/a, and 800 Gg/a. Township level
residue yield (Mg ha^-1^), cycle times (h), and harvest areas (ha) are
the three essential summary parameters used by the greedy optimization
model.

## Data aggregation

### Cycle time estimates

@@

$$\bar{s_{h}} = \left( \frac{s_{l}^{- 1} + s_{u}^{- 1}}{2} \right)^{- 1}$$

Where $s_{l}$ is the loaded speed, and $s_{u}$ is the unloaded speed.
Use of harmonic mean allows us to calculate the correct cycle time: the
time required to travel from the mill to the collection site at the
cutblock, and back to the mill. This provides the basis for our
estimates of haul cost.

Figure 2 illustrates the road network in Alberta and estimates of cycle
time based on the average of loaded and unloaded speeds from Table 1.
The road network in the White Area of the province is very dense and
largely arranged as a grid defined by the Alberta Township Survey
System. Most of the roads in the White Area are public. The road network
in the Green Area is less dense and many of the roads are built and
maintained by resource industries (*e.g.,* forestry, petroleum, and
natural gas). The pattern of the road network is not as regular as in
the White Area.

[ Figure 2 about here ]

One-way haul times based on the harmonic mean travel speed were
generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
@Raffler2018, specifically its iso-area as interpolation (from point)
algorithm. In our case, we used the algorithm to determine the minimum
one-way travel time (using the harmonic mean of loaded and unloaded
speeds) along the road network from each of our centers to each cell of
a provincial level raster at a 200 m resolution. Off-road travel was
assumed to be at 5 km·h^-1^. The Dijkstra algorithm @Dijkstra1959 is the basis
for travel time minimization in QNEAT3. Cycle time is double the one-way
haul time with an extra hour added for loading and unloading.

QGIS zonal statistics were used to calculate the median cycle time to
each township, from each of the three potential bioenergy plants. The
cycle time in Figure 2 was limited to 14 hours because, in Alberta,
drivers of commercial vehicles are not permitted to drive more than 13
hours in a 24-hour period and can spend no more than 14 hours on-duty,
including loading and unloading times.

### Biomass collection and transportation costs

All costs and prices used in this study are in Canadian dollars (\$). At
the time of writing (2021-11-05), the exchange rate between Canadian
(CAD) and US dollars (USD) was 0.8031 USD/CAD.

We assume that the biomass available for collection at a cutblock
includes the tops, branches, and foliage left at a roadside landing
after processing by a stroke delimber. The delimber is assumed to cut
the top of the log off at the diameter specified in the utilization
standard specific to the harvest area. Following Ref. @Roser2013, the cost
of collecting residues left behind by a roadside stroke delimber, which
includes pre-piling, chipping, road maintenance, supervision, and
loading (but does not consider transport) is 41.60 \$·Mg^-1^.
